Motion planning for assessing accessibility of protein tunnels

### 2017-09-09 paper.tex
graphs for comparing methods a1,a2, ligand images, expeirments

Therefore, we propose to measure the distance as the smallest 3D Euclidean distance between atoms placed at $\qrand$ and a configuration
$q$ being tested.
This metric is denoted $\da(\cdot,\cdot)$ on line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}.
As the expansion routine attempts to find such sample that minimizes this distance to $\qrand$, it supports retraction of the ligand towards $\qrand$.
The fast evaluation of this metric with complexity $O(\log n)$, where $n$ is the number of atoms of the ligand, can be achieved using KD-trees.
For each conformation $\l \in L$ of the ligand, a KD-tree is build for the ligand at configuration $q=(0,0,0,0,0,0,l,\smax)$
(i.e., for ligand with no translation and no rotation).
To find a nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $T=(x,y,z)$ and
rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-T)$ and the KD-tree search is used to find nearest point towards $\qrand'$, which
is also the nearest neighbor of the ligand at $q$ to $\qrand$.

@@

\begin{figure}
\centering
\includegraphics[width=0.3\textwidth]{fig/m040-conformations}
\caption{\label{fig::m040c}
 Examples of `long' conformation (left) and a `packed' one for the ligand M040 (1,5-Dichloropentane).
}
\end{figure}

@@

{\def\a{270}
\begin{figure}
\centering
\renewcommand{\tabcolsep}{0pt}
\begin{tabular}{cccc}
\includegraphics[width=0.15\textwidth,angle=\a]{fig/m001-oneAtom-cdCount} &
\includegraphics[width=0.15\textwidth,angle=\a]{fig/m001-oneAtom-rrt_path_time} &
\includegraphics[width=0.15\textwidth,angle=\a]{fig/m037t-cdCount} & 
\includegraphics[width=0.15\textwidth,angle=\a]{fig/m037t-rrt_path_time} \\
\multicolumn{2}{c}{One-atom} &
\multicolumn{2}{c}{m037}  \\
\multicolumn{4}{c}{%
\includegraphics[width=0.15\textwidth,angle=\a]{fig/m040-10-cdCount} 
\includegraphics[width=0.15\textwidth,angle=\a]{fig/m040-10-rrt_path_time} 
\includegraphics[width=0.15\textwidth,angle=\a]{fig/m040-10-trajDistToGoal}} \\
\multicolumn{4}{c}{m040} 
\end{tabular}
\end{figure}
}

@@

The aim of this experiment is to verify the procedure to select ligand conformations for the motion planning.
The experiment was performed with the ligand DCP, for which $1300$ conformations where available and for the ligand
M040 with $7560$ available conformations.
Two strategies were tested: the strategy described in Sec.~\ref{sec::strat} (denoted `S') and a strategy,
that prefers smaller conformations (denoted `P').
In the second method, deviation of the atomic distance from the geometric center are computed and the conformations
with the lowest deviations are selected.
The results are summarized in Tab.~\ref{tabl::selection}, which shows the percentage number of frames that are traversable by the tested ligand.
For each tested ligand, 5 and 50 conformations from the available pool of conformations were selected using the strategies.
In the case of both tested ligands and 50 selected conformations, the `S' strategy leads to higher traversability rate than the `P' strategy.
For the ligand DCP, a higher traversability rates was achieved when 50 conformations are selected.
This experiments shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
than using less number of conformations or when preferring the packed conformations.

\begin{table}
\centering
\caption{\label{tab::selection}Traversability of the tunnels depending on conformation selection}
\small
\renewcommand{\tabcolsep}{3pt}
{\small
\input graphSel.tex
}
\end{table}

@@

The aim of the second experiment is to determine which ligands are

The results are shown in Tab.~\ref{tab::main}.

\begin{table}
\centering
\caption{\label{tab::main}Traversability of the tunnels}
\small
\renewcommand{\tabcolsep}{3pt}
{\small
\input graph.tex
}
\end{table}

>>>>

Therefore, we propose to measure the distance as the smallest 3D Euclidean distance between atoms placed at $\qrand$ and a configuration
$q$ being tested.
This metric is denoted $\da(\cdot,\cdot)$ on line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}.
As the expansion routine attempts to find such sample that minimizes this distance to $\qrand$, it supports retraction of the ligand towards $\qrand$.
The fast evaluation of this metric with complexity $O(\log n)$, where $n$ is the number of atoms of the ligand, can be achieved using KD-trees.
For each conformation $\l \in \L$ of the ligand, a KD-tree is build for the ligand at configuration $q=(0,0,0,0,0,0,l,\smax)$
(i.e., for ligand with no translation and no rotation).
To find a nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $T=(x,y,z)$ and
rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-T)$ and the KD-tree search is used to find nearest point towards $\qrand'$, which
is also the nearest neighbor of the ligand at $q$ to $\qrand$.

@@

\begin{figure}
\centering
%\includegraphics[width=0.25\textwidth]{fig/m040-conformations}
\includegraphics[width=0.15\textwidth]{fig/m040-conf1} \hskip 25pt
\includegraphics[width=0.12\textwidth]{fig/m040-conf2}
\caption{\label{fig::m040c}
 Examples of `long' conformation (left) and a `packed' one for the ligand m040 (1,5-Dichloropentane).
}
\end{figure}

@@

%
%{\def\a{270}
%\begin{figure}
%\centering
%\renewcommand{\tabcolsep}{0pt}
%\begin{tabular}{cccc}
%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m001-oneAtom-cdCount} &
%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m001-oneAtom-rrt_path_time} &
%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m037t-cdCount} & 
%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m037t-rrt_path_time} \\
%\multicolumn{2}{c}{One-atom} &
%\multicolumn{2}{c}{m037}  \\
%\multicolumn{4}{c}{%
%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m040-10-cdCount} 
%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m040-10-rrt_path_time} 
%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m040-10-trajDistToGoal}} \\
%\multicolumn{4}{c}{m040} 
%\end{tabular}
%\end{figure}
%}
%

@@

The aim of this experiment is to verify the procedure to select ligand conformations for the motion planning.
The experiment was performed with the ligand DCP (Dichloropropanol), for which $1300$ conformations where available and for the ligand
m040 (1,5-Dichloropentane) with $7560$ available conformations.
Two strategies were tested: the strategy described in Sec.~\ref{sec::strat} (denoted `S') and a strategy that prefers smaller conformations (denoted `P').
The `P' method selects such conformations where the deviation of the atomic distance from the geometric center are lowest, which leads 
to selection of the packed conformations (Fig.~\ref{m040c}).

The percentage of frames that were traversible by the ligands are shown in Tab.~\ref{tab::selection}, where the subscript of each ligand
name shows the number of selected conformations (5 or 50) and the used method (S or P).
The number after the slash in the ligand's name denote the number of atoms of the ligand.
In the case of both tested ligands and 50 selected conformations, the `S' strategy leads to higher traversability rate than the `P' strategy.
For the ligand DCP, a higher traversability rates was achieved when 50 conformations are selected.
This experiments shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
than using less number of conformations or when preferring the packed conformations.

\begin{table}
\centering
\caption{\label{tab::selection}
    Traversability of the tunnels depending on conformation selection.
    The number after $/$ denotes the number of atoms.
}
\small
\renewcommand{\tabcolsep}{3pt}
{\small
\input graphSel.tex
}
\end{table}

@@

Based on the results of the previous experiment, 50 conformations using the `S' strategy were selected for the following
ligands: m037t (1,2-Dichloroethane), m038t (1,3-Dichloropropane), m056 ( 2-Chlorobutane) and m080 (Trichloropropane), and
the task was to compute trajectories for these ligands in the MD sequence.

The results are summarized in Tab.~\ref{tab::main}, which shows percentage of frames in which the tested planners \RA\ and \RB\ found
trajecotries close to the active site. 
The active site was considered as reached if the distance from the center of the ligand to the active site was less than 2~\AA.

The analyzes sequence of MD  snapshots contains two tunnels. 
The average bottleneck of the tunnels are $1.48$~\AA and $1.2$~\AA for the first, resp. the second tunnel.
The tables shows percentages of traversable MD snapshots only for the first tunnel, as no trajectories were found in the second tunnel
for most of the tested ligand.
The only except is the ligand m037t, which was the smallest tested ligand (with 8 atoms), and therefore it can fit even inside the second tunnel.

The proposed \RA\ planner outperforms the \RB\ planner in the number of frames where a solution was found.
The distances of the configuration trees (measured using 3D Euclidean metric) to the active site and the runtimes in each frame
are shown in Fig.~\ref{fig::compared}.
The runtimes of both algorithms increases in those frames, where the tunnel is not traversable, i.e., where the distance
to the active site is more than 2~\AA (e.g., the frame number 60).

\begin{table}
\centering
\caption{\label{tab::main}
    Traversability of the tunnels for ligands with 50 conformations. 
    The number after $/$ denotes the number of atoms.
}
\small
\renewcommand{\tabcolsep}{3pt}
{\small
\input graph.tex
}
\end{table}

{\def\a{270}
\begin{figure}
\centering
\includegraphics[width=0.25\textwidth,angle=\a]{fig/m037tmins-03tunnel-1rrt_path_time-2}
\includegraphics[width=0.25\textwidth,angle=\a]{fig/m037tmins-03tunnel-1trajDistToGoal-2}
\caption{\label{fig::comparison}
    Runtimes and distance to the active site for the m037 ligand at scale 0.4.
    The values are computed from all 200 trials (20 initial configurations $\times$ 10 trials).
    Only each second frame is shown in the graph.
}
\end{figure}
}
